---
title: "Combine Scraped Data"
output: md_document
category: science
---

```{r setup, include=FALSE} 
# Figure path on disk = base.dir + fig.path
# Figure URL online = base.url + fig.path
knitr::opts_knit$set(base.dir = "/Users/Shared/AttM/", base.url="/AttM/")
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = "rmd_images/2021-01-01-combine_scraped_data/")

# knitr::opts_chunk$set(warning = FALSE, message = FALSE, root.dir = "/Users/Shared/AttM") 
# setwd("/Users/Shared/AttM")
```

 <br>
   
## Description

This script combines fight data that has been scraped from various sites.  

 <br>
   
### Libraries 

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

 <br>
     
### Combine Databases Events & Odds   

 <br>
   
#### Pre-process

Load data.

```{r}
load("./Datasets/fight_odds.RData")
load("./Datasets/event_info.RData")
```

Ensure that dates are characters. 

```{r}
event_info$Date = as.character(event_info$Date)
```

Add loser column. 

```{r}
fight_odds %>%
  dplyr::mutate(Loser = ifelse(Winner == Fighter1, Fighter2, Fighter1)) -> fight_odds
```

Identify odds by Winner/Loser label.

```{r}
fight_odds %>%
  dplyr::mutate(
    Winner_Odds = ifelse(Winner == Fighter1, Fighter1_Decimal_Odds, Fighter2_Decimal_Odds)
    , Loser_Odds = ifelse(Loser == Fighter1, Fighter1_Decimal_Odds, Fighter2_Decimal_Odds)
  ) -> fight_odds
```

Only keep columns of interest. 

```{r}
fight_odds %>% dplyr::select(
  -c(
    "Events"
    , "Fighter1"
    , "Fighter1_Decimal_Odds"
    , "Fighter2"
    , "Fighter2_Decimal_Odds"
    )
  ) -> fight_odds_clean
```

Identify date discrepancies as odds website appears to have some of the dates wrong if compared to UFC website.   

```{r}
setdiff(fight_odds_clean$Date, event_info$Date)
```

Correct manually for these discrepancies

```{r}
fight_odds_clean$Date = dplyr::recode(
  fight_odds_clean$Date
  , "2020-12-06" = "2020-12-05"
  , "2020-11-29" = "2020-11-28"
  , "2020-10-04" = "2020-10-03"
  , "2020-05-14" = "2020-05-13"
  , "2020-05-10" = "2020-05-09"
  , "2020-03-08" = "2020-03-07"
  , "2020-02-09" = "2020-02-08"
  , "2019-12-15" = "2019-12-14"
  , "2018-02-04" = "2018-02-03"
  , "2017-12-31" = "2017-12-30"
  , "2017-12-02" = "2017-12-01"
  , "2016-12-31" = "2016-12-30"
  , "2013-12-07" = "2013-12-06"
)
```

Get rid of either the Loser or Winner column to merge below. 

```{r}
fight_odds_clean %>% dplyr::select(
  -c("Loser")
) -> fight_odds_clean_win
  
fight_odds_clean %>% dplyr::select(
  -c("Winner")
) -> fight_odds_clean_lose
```

Merge by (either Winner or Loser) and Date

```{r}
df_fight_win = merge(event_info, fight_odds_clean_win, by = c("Winner", "Date"))
df_fight_lose = merge(event_info, fight_odds_clean_lose, by = c("Loser", "Date"))
```

Add these databases together.

```{r}
df_fight_bind = rbind(df_fight_win, df_fight_lose)
```

Remove duplicates.

```{r}
df_fight_bind %>% distinct() -> event_and_odds 
```

 <br>
    
#### Quality Control

How much data (# rows) do we lose if compared to fight_odds_clean data frame?  

```{r}
nrow(fight_odds_clean) - nrow(event_and_odds )   
```

Which fight dates, if any, did we lose entirely?  

```{r}
lost_fight_dates = unique(fight_odds_clean$Date)[!(unique(fight_odds_clean$Date) %in% unique(event_and_odds$Date))]
```

Which events do these dates correspond to?

```{r}
event_info %>%
  dplyr::filter(Date %in% lost_fight_dates) %>%
  group_by(Date) %>%
  dplyr::summarise(Events = unique(Event))
```

Which Winners and Losers were lost during merging?

```{r}
lost_winners = unique(fight_odds_clean$Winner)[!(unique(fight_odds_clean$Winner) %in% unique(event_and_odds$Winner))]
lost_losers = unique(fight_odds_clean$Loser)[!(unique(fight_odds_clean$Loser) %in% unique(event_and_odds$Loser))]
```

What other data did we lose (based on Odds and Date)?  

The below code identifies the discrepancies between the fight_oods_clean and event_and_odds dataframes.  

```{r}
fight_odds_clean %>% 
  dplyr::select(
    Date
    , Winner_Odds
    , Loser_Odds
  ) -> fight_odds_clean_odds

event_and_odds %>%
  dplyr::select(
    Date
    , Winner_Odds
    , Loser_Odds
  ) -> event_and_odds_odds

fight_odds_clean_odds$Coder = "b"
event_and_odds_odds$ Coder = "d"

df_compare = rbind(event_and_odds_odds, fight_odds_clean_odds)

dupsBetweenGroups <- function (df, idcol) {
  # df: the data frame
  # idcol: the column which identifies the group each row belongs to
  
  # Get the data columns to use for finding matches
  datacols <- setdiff(names(df), idcol)
  
  # Sort by idcol, then datacols. Save order so we can undo the sorting later.
  sortorder <- do.call(order, df)
  df <- df[sortorder,]
  
  # Find duplicates within each id group (first copy not marked)
  dupWithin <- duplicated(df)
  
  # With duplicates within each group filtered out, find duplicates between groups. 
  # Need to scan up and down with duplicated() because first copy is not marked.
  dupBetween = rep(NA, nrow(df))
  dupBetween[!dupWithin] <- duplicated(df[!dupWithin,datacols])
  dupBetween[!dupWithin] <- duplicated(df[!dupWithin,datacols], fromLast=TRUE) | dupBetween[!dupWithin]
  
  # ============= Replace NA's with previous non-NA value ==============
  # This is why we sorted earlier - it was necessary to do this part efficiently
  
  # Get indexes of non-NA's
  goodIdx <- !is.na(dupBetween)
  
  # These are the non-NA values from x only
  # Add a leading NA for later use when we index into this vector
  goodVals <- c(NA, dupBetween[goodIdx])
  
  # Fill the indices of the output vector with the indices pulled from
  # these offsets of goodVals. Add 1 to avoid indexing to zero.
  fillIdx <- cumsum(goodIdx)+1
  
  # The original vector, now with gaps filled
  dupBetween <- goodVals[fillIdx]
  
  # Undo the original sort
  dupBetween[sortorder] <- dupBetween
  
  # Return the vector of which entries are duplicated across groups
  return(dupBetween)
}

dupRows <- dupsBetweenGroups(df_compare, "Coder")

df_dups = cbind(df_compare, dupRows)

lost_fight_odds = df_dups[df_dups$dupRows == F,]

lost_fights = merge(lost_fight_odds, fight_odds_clean)
```

The following fight data were lost.  

Unsure of exact reason for these lost fights but finding out could be tedious. They could be cases where both fight names were mismatches.  

```{r}
lost_fights[,c(1,6,7)]
```

 <br>
    
#### Save Data

```{r}
save(event_and_odds, file = "./Datasets/event_and_odds.RData")
```

 <br>
     
### Add Fighter Stats Database

NOTE: the following fighter stats are dated to when the data were scrapped from the UFC website. Therefore, once merged with the events and odds database, the stats will not necessarily be representative of what they were the night a given fighter was fighting.   

 <br>
   
#### Pre-process   
  
Load data:

```{r}
load("./Datasets/fighter_stats.RData")
```

Redefine weight classes.

```{r}
fighter_stats$WeightClass = dplyr::recode(
  fighter_stats$WeightClass
  , "lightweight" = "Lightweight"
  , "heavyweight" = "Heavyweight"
  , "featherweight" = "Featherweight"
  , "welterweight" = "Welterweight"
  , "middleweight" = "Middleweight"
  , "lightheavyweight" = "Light Heavyweight"
  , "catchweight" = "Catch Weight"
  , "flyweight" = "Flyweight"
  , "strawweight" = "Strawweight"
  , "bantamweight" = "Bantamweight"
)
```

Rename columns. 

```{r}
fighter_stats %>%
  dplyr::rename(
    "FighterWeightClass" = "WeightClass"
    , "FighterWeight" = "Weight"
  ) -> fighter_stats

event_and_odds %>%
  dplyr::rename(
    "FightWeightClass" = "WeightClass"
  ) -> event_and_odds
```

Add Sex category.

```{r}
event_and_odds %>%
  dplyr::mutate(
    Sex = ifelse(grepl("Women's ", FightWeightClass), "Female", "Male")
  ) -> event_and_odds
```

Remove mention of Sex from Weight Class category.

```{r}
event_and_odds$FightWeightClass = gsub("Women's ", "", event_and_odds$FightWeightClass)
```

Check that Weight Class labels make sense. 

```{r}
unique(event_and_odds$FightWeightClass)
```

Create unique fight id for each fight (i.e. each row).

```{r}
event_and_odds$fight_id = 1:nrow(event_and_odds)
```

Gather data frame by fighter (i.e. go from short to long format).

```{r}
event_and_odds %>%
  gather(Result, NAME, c("Winner", "Loser")) -> event_and_odds_long
```

Merge databases.

```{r}
df_master = merge(event_and_odds_long, fighter_stats, by = c("NAME"))
```

 <br>
     
#### Quality Control

Examine difference in data frame length between df_master and event_and_odds long. 

```{r}
nrow(df_master) - nrow(event_and_odds_long)
```

Are there any duplicate rows in df_master?

```{r}
df_master %>% distinct() -> df_master
```

Where are the additional rows coming from?

```{r}
sum(duplicated(fighter_stats$NAME))
```

Who are these duplicate fighters? 

```{r}
fighter_stats[duplicated(fighter_stats$NAME),]
```

What are the duplicate entries?

```{r}
fighter_stats[fighter_stats$NAME %in% fighter_stats$NAME[duplicated(fighter_stats$NAME)], ]
```

Remove these entries if they have Weight Class mismatches. 

```{r}
dup_names = fighter_stats$NAME[duplicated(fighter_stats$NAME)]

df_master %>%
  dplyr::filter(
    !((NAME %in% dup_names) & (FighterWeightClass != FightWeightClass))
  ) -> df_master 
```

Remove fight IDs without pair.

```{r}
df_master %>%
  group_by(fight_id) %>%
  summarize(count = length(Sex)) %>%
  dplyr::filter(count != 2) -> to_remove_from_master

df_master[df_master$fight_id %in% to_remove_from_master$fight_id,]

df_master = df_master[!(df_master$fight_id %in% to_remove_from_master$fight_id),]
```

 <br>
   
#### Save File

```{r}
save(df_master, file = "./Datasets/df_master.RData")
```

 <br>
     
### Instead, Add CURRENT Fighter Stats Database

NOTE: unlike the previously fighter stats database that was added, these fighter stats are as they were at the time of the actual UFC events (hence "current"). This current fighter stats database will be added to the events and odds database in the same manner as the previous fighter stats database.  

 <br>
     
#### Pre-process

Load data.

```{r}
load("./Datasets/current_fighter_stats.RData")
```

Redefine weight classes.

```{r}
current_fighter_stats$WeightClass = dplyr::recode(
  current_fighter_stats$WeightClass
  , "lightweight" = "Lightweight"
  , "heavyweight" = "Heavyweight"
  , "featherweight" = "Featherweight"
  , "welterweight" = "Welterweight"
  , "middleweight" = "Middleweight"
  , "lightheavyweight" = "Light Heavyweight"
  , "catchweight" = "Catch Weight"
  , "flyweight" = "Flyweight"
  , "strawweight" = "Strawweight"
  , "bantamweight" = "Bantamweight"
)
```

Rename columns.

```{r}
current_fighter_stats %>%
  dplyr::rename(
    "FighterWeightClass" = "WeightClass"
    , "FighterWeight" = "Weight"
  ) -> current_fighter_stats

current_fighter_stats %>%
  dplyr::rename(
    "Date" = "Event_Date" 
  ) -> current_fighter_stats
```

Convert Date to character type. 

```{r}
current_fighter_stats$Date = as.character(current_fighter_stats$Date)
```

Merge current fighter database with events and odds database. 

```{r}
df_current_master = merge(event_and_odds_long, current_fighter_stats, by = c("NAME", "Date"))
```

Get distinct entries.

```{r}
df_current_master %>% distinct() -> df_current_master
```

Get rid of fight IDs without pair. 

```{r}
df_current_master %>%
  group_by(fight_id) %>%
  summarize(count = length(Sex)) %>%
  dplyr::filter(count != 2) -> to_remove_from_master

df_current_master[df_current_master$fight_id %in% to_remove_from_master$fight_id,]

df_current_master = df_current_master[!(df_current_master$fight_id %in% to_remove_from_master$fight_id),]
```

 <br>
     
#### Save File 

```{r}
save(df_current_master, file = "./Datasets/df_current_master.RData")
```

