---
title: "Get Fighter Stats"
output: md_document
category: science
---

```{r setup, include=FALSE} 
# Figure path on disk = base.dir + fig.path
# Figure URL online = base.url + fig.path
knitr::opts_knit$set(base.dir = "/Users/Shared/AttM/", base.url="/AttM/")
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = "rmd_images/2021-01-01-get_fighter_stats/")

# knitr::opts_chunk$set(warning = FALSE, message = FALSE, root.dir = "/Users/Shared/AttM") 
# setwd("/Users/Shared/AttM")
```

## Description

This script scrapes UFC fighter information from the http://www.fightmetric.com website.

The script is adapted from:

https://github.com/jasonchanhku/UFC-MMA-Predictor/blob/master/UFC%20MMA%20Predictor%20Workflow.ipynb

 <br>
   
### Libraries

```{r}
library(rvest)
library(dplyr)
library(tidyr)
library(stringr)
```

 <br>
   
### Web Scrape Fighter Stats

Loop through alphabet to get list of all desired webpages. 

```{r}
webpage <- character(length(letters))
url_list <- c()

for(j in 1:length(letters)){
  
  webpage[j] <- paste0("http://www.fightmetric.com/statistics/fighters?char=", letters[j], "&page=all")
  temp <- read_html(webpage[j]) %>% html_nodes(".b-statistics__table-col:nth-child(1) .b-link_style_black") %>% html_attr("href")
  url_list <- c(url_list, temp)     
}
```

 <br>
   
#### Initialization

Create data structures to store the following fighter stats: 

LEGEND:
SLpM - Significant Strikes Landed per Minute
Str. Acc. - Significant Striking Accuracy
SApM - Significant Strikes Absorbed per Minute
Str. Def. - Significant Strike Defence (the % of opponents strikes that did not land)
TD Avg. - Average Takedowns Landed per 15 minutes
TD Acc. - Takedown Accuracy
TD Def. - Takedown Defense (the % of opponents TD attempts that did not land)
Sub. Avg. - Average Submissions Attempted per 15 minutes

```{r}
name <- character(length(url_list))
weightclass <- numeric(length(url_list))
reach <- numeric(length(url_list))
slpm <- numeric(length(url_list))
td <- numeric(length(url_list))
tda <- numeric(length(url_list))
tdd <- numeric(length(url_list))
stra <- numeric(length(url_list))
strd <- numeric(length(url_list))
suba <- numeric(length(url_list))
sapm <- numeric(length(url_list))
```

 <br>
   
#### Loop

```{r}

# start the clock
ptm <- proc.time()

for(i in 1:length(url_list)) {
  
  # # print iteration
  # print(sprintf("%d of %d", i, length(url_list)))
  # # start the clock
  # ptm <- proc.time()
  
  # fighter url
  fighter_url <- read_html(url_list[i])
  
  # fighter name
  name_t <- fighter_url %>% html_nodes(".b-content__title-highlight") %>% html_text()
  name_t <- gsub("\n", "", name_t)
  name[i] <- as.character(trimws(name_t))
  
  # weightclass
  weightclass_t <- fighter_url %>% html_nodes(".b-list__info-box_style_small-width .b-list__box-list-item_type_block:nth-child(2)") %>% html_text()
  weightclass_t <- gsub(" ", "", weightclass_t)
  weightclass_t <- gsub("\n", "", weightclass_t)
  weightclass_t <- strsplit(weightclass_t, ":")
  weightclass_t <- weightclass_t[[1]][2]
  weightclass_t <- strsplit(weightclass_t, "l")
  weightclass[i] <- as.numeric(as.character(weightclass_t[[1]][1]))
  
  # reach
  reach_t <- fighter_url %>% html_nodes(".b-list__info-box_style_small-width .b-list__box-list-item_type_block:nth-child(3)") %>% html_text()
  reach_t <- gsub(" ", "", reach_t)
  reach_t <- gsub("\n", "", reach_t)
  reach_t <- strsplit(reach_t, ":")
  reach_t <- reach_t[[1]][2]
  reach_t <- strsplit(reach_t, "\"")
  reach[i] <- as.numeric(as.character(reach_t[[1]][1]))
  
  # feature 1: slpm
  slpm_t <- fighter_url %>% html_nodes(".b-list__info-box-left .b-list__info-box-left .b-list__box-list-item_type_block:nth-child(1)") %>% html_text()
  slpm_t <- gsub(" ", "", slpm_t)
  slpm_t <- gsub("\n", "", slpm_t)
  slpm_t <- strsplit(slpm_t, ":")
  slpm_t <- slpm_t[[1]][2]
  slpm[i] <- as.numeric(slpm_t)
  
  # feature 2: takedown avg
  td_t <- fighter_url %>% html_nodes(".b-list__info-box_style-margin-right .b-list__box-list-item_type_block:nth-child(2)") %>% html_text()
  td_t<- gsub(" ", "", td_t)
  td_t <- gsub("\n", "", td_t)
  td_t <- strsplit(td_t, ":")
  td_t <- td_t[[1]][2]
  td[i] <- as.numeric(td_t)
  
  # feature 3: significant striking accuracy
  stra_t <- fighter_url %>% html_nodes(".b-list__info-box-left .b-list__info-box-left .b-list__box-list-item_type_block:nth-child(2)") %>% html_text()
  stra_t <- gsub(" ", "", stra_t)
  stra_t <- gsub("\n", "", stra_t)
  stra_t <- strsplit(stra_t, ":")
  stra_t <- stra_t[[1]][2]
  stra_t <- strsplit(stra_t, "%")
  stra[i] <- as.numeric(stra_t) / 100
  
  # feature 4: takedown accuracy 
  tda_t <- fighter_url %>% html_nodes(".b-list__info-box_style-margin-right .b-list__box-list-item_type_block:nth-child(3)") %>% html_text()
  tda_t <- gsub(" ", "", tda_t)
  tda_t <- gsub("\n", "", tda_t)
  tda_t <- strsplit(tda_t, ":")
  tda_t <- tda_t[[1]][2]
  tda_t <- strsplit(tda_t, "%")
  tda[i] <- as.numeric(tda_t) / 100
  
  # feature 5: significant absorbed stras per minute
  sapm_t <- fighter_url %>% html_nodes(".b-list__info-box-left .b-list__info-box-left .b-list__box-list-item_type_block:nth-child(3)") %>% html_text()
  sapm_t <- gsub(" ", "", sapm_t)
  sapm_t <- gsub("\n", "", sapm_t)
  sapm_t <- strsplit(sapm_t, ":")
  sapm_t <- sapm_t[[1]][2]
  sapm[i] <- as.numeric(sapm_t)
  
  # feature 6: takedown def
  tdd_t <- fighter_url %>% html_nodes(".b-list__info-box_style-margin-right .b-list__box-list-item_type_block:nth-child(4)") %>% html_text()
  tdd_t <- gsub(" ", "", tdd_t)
  tdd_t <- gsub("\n", "", tdd_t)
  tdd_t <- strsplit(tdd_t, ":")
  tdd_t <- tdd_t[[1]][2]
  tdd_t <- strsplit(tdd_t, "%")
  tdd[i] <- as.numeric(tdd_t) / 100
  
  # feature 7: striking def
  strd_t <- fighter_url %>% html_nodes(".b-list__info-box-left .b-list__info-box-left .b-list__box-list-item_type_block:nth-child(4)") %>% html_text()
  strd_t <- gsub(" ", "", strd_t)
  strd_t <- gsub("\n", "", strd_t)
  strd_t <- strsplit(strd_t, ":")
  strd_t <- strd_t[[1]][2]
  strd_t <- strsplit(strd_t, "%")
  strd[i] <- as.numeric(strd_t) / 100
  
  # feature 8: submission average
  suba_t <- fighter_url %>% html_nodes(".b-list__box-list_margin-top .b-list__box-list-item_type_block:nth-child(5)") %>% html_text()
  suba_t <- gsub(" ", "", suba_t)
  suba_t <- gsub("\n", "", suba_t)
  suba_t <- strsplit(suba_t, ":")
  suba_t <- suba_t[[1]][2]
  suba[i] <- as.numeric(suba_t)
  
  # # stop the clock
  # print(proc.time() - ptm) 
}

# stop the clock
print(proc.time() - ptm) 
```

 <br>
   
### Manipulate Data Frame

Create data frame and add column describing weightclass. 

```{r}
database <- data.frame(NAME = name, Weight = weightclass, REACH = reach, SLPM = slpm, SAPM = sapm, STRA = stra, STRD = strd, TD = td, TDA = tda, TDD = tdd, SUBA = suba )

data_add_wc <- mutate(database, WeightClass = ifelse(Weight == 115, "strawweight", ifelse(Weight == 125, "flyweight", ifelse(Weight == 135, "bantamweight", ifelse(Weight == 145, "featherweight", ifelse(Weight == 155, "lightweight", ifelse(Weight == 170, "welterweight", ifelse(Weight == 185, "middleweight", ifelse(Weight == 205, "lightheavyweight", ifelse(Weight > 205, "heavyweight", "catchweight"))))))))))

data_add_wc <- data_add_wc[c(1,2,12,3,4,5,6,7,8,9,10,11)]
```

The following fighters did not have their weight listed. 

```{r}
data_add_wc[is.na(data_add_wc$Weight), ]
```

Get rid of fighters with no weight listed. 

```{r}
fighter_stats <- data_add_wc[!is.na(data_add_wc$Weight), ]
```

The following fighters are listed at catchweight. 

```{r}
fighter_stats %>% filter(WeightClass == "catchweight")
```

Examine data frame.

```{r}
summary(fighter_stats)
```

 <br>
   
### Save data.

```{r}
save(fighter_stats, file ="./Datasets/fighter_stats.RData")
```

