---
title: "Over Rated Fighters"
author: "Aggression to the Mean"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: md_document
category: arts
---

```{r setup, include=FALSE}
# Figure path on disk = base.dir + fig.path
# Figure URL online = base.url + fig.path
knitr::opts_knit$set(base.dir = "/Users/Shared/AttM/", base.url="/AttM/")
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.path = "rmd_images/2021-02-19-over_rated_fighters/")

# ------

# knitr::opts_chunk$set(warning = FALSE, message = FALSE, root.dir = "/Users/Shared/AttM")
# setwd("/Users/Shared/AttM")
```


 <br>

## Introduction

In [another post](https://aggression-to-the-mean.github.io/AttM/for-the-people/arts/2021/02/19/Under-Rated-Fighters.html), we identified some of the UFC's most Under Rated fighters by examining the extent to which they had overperformed relative to the odds.  

The purpose of this post is to identify some of the most Over Rated fighters in the UFC.   

<br> 

# Defining "Over Rated"

In another post, to assess the extent to which a fighter was Under Rated, we examined if the fighter won at a greater rate than the average Implied Probability of their odds (i.e. has the fighter been overperforming)?

In this post, we will examine the converse: has a fighter **lost** at a greater rate than the average  Implied Probability of their odds (i.e. has the fighter been underperforming). Indeed, if a fighter has been underperforming, then they have been effectively Over Rated by the odds. 

<br> 

# Dataset - Historical Data of UFC Fight Odds 

```{r include = F, echo = F}
# packages
library(tidyverse)
library(knitr)

## ------------------------------------------------------------------------------------------------------
load("./Datasets/df_master.RData")


## ------------------------------------------------------------------------------------------------------
summary(df_master)


## ------------------------------------------------------------------------------------------------------
df_master$NAME = as.factor(df_master$NAME)
df_master$Date = as.Date(df_master$Date)
df_master$Event = as.factor(df_master$Event)
df_master$City= as.factor(df_master$City)
df_master$State = as.factor(df_master$State)
df_master$Country = as.factor(df_master$Country)
df_master$FightWeightClass = as.factor(df_master$FightWeightClass)
df_master$Method = as.factor(df_master$Method)
df_master$Winner_Odds = as.numeric(df_master$Winner_Odds)
df_master$Loser_Odds = as.numeric(df_master$Loser_Odds)
df_master$fight_id = as.factor(df_master$fight_id)
df_master$Sex = as.factor(df_master$Sex)
df_master$Result = as.factor(df_master$Result)
df_master$FighterWeightClass = as.factor(df_master$FighterWeightClass)

## ------------------------------------------------------------------------------------------------------
df_odds = df_master
rm(df_master)


## ------------------------------------------------------------------------------------------------------
df_odds %>%
  dplyr::filter(
    (Method != "DQ") & (Method != "Overturned")
    , is.finite(Winner_Odds)
    , is.finite(Loser_Odds)
  ) -> df_odds


## ------------------------------------------------------------------------------------------------------
df_odds %>%
  dplyr::select(-c(FighterWeight:SUBA)) %>%
  spread(Result, NAME) -> df_odds_short

## ------------------------------------------------------------------------------------------------------
df_odds_short %>%
  dplyr::filter(Winner_Odds != Loser_Odds) %>%  # filter out equal odds
  dplyr::mutate(
    Favorite_was_Winner = ifelse(Winner_Odds < Loser_Odds, T, F)
    , Favorite_Unit_Profit = ifelse(Favorite_was_Winner, Winner_Odds - 1, -1)
    , Underdog_Unit_Profit = ifelse(!Favorite_was_Winner, Winner_Odds - 1, -1)
  ) -> df_odds_short

## ------------------------------------------------------------------------------------------------------
df_odds_short %>% dplyr::mutate(
  Favorite_Probability = ifelse(Favorite_was_Winner, 1/Winner_Odds, 1/Loser_Odds)
  , Underdog_Probability = ifelse(!Favorite_was_Winner,  1/Winner_Odds, 1/Loser_Odds)
) -> df_odds_short


## ------------------------------------------------------------------------------------------------------
df_odds_short %>%
  dplyr::mutate(
    Total_Probability = Favorite_Probability + Underdog_Probability
    , Overround = Total_Probability - 1
  ) -> df_odds_short

df_odds_short %>%
  dplyr::mutate(
    Adjusted_Favorite_Probability = Favorite_Probability - Overround/2
    , Adjusted_Underdog_Probability = Underdog_Probability - Overround/2
    , Adjusted_Total_Probability = Adjusted_Favorite_Probability + Adjusted_Underdog_Probability
  ) -> df_odds_short

```

Using data science tools such as web scraping, I have obtained a dataset with fight odds information regarding the majority of the UFC fights that occurred between 2013 and present.     

```{r include = F, echo = F}
nfights = nrow(df_odds_short)
nevents = length(unique(df_odds_short$Event))
earliest_date = format(min(df_odds_short$Date), '%B %d, %Y')
latest_date = format(max(df_odds_short$Date), '%B %d, %Y')
```

In particular, the dataset consists of `r nfights` UFC fights from `r nevents` UFC events, spanning from `r earliest_date` to `r latest_date`.  

Among other things, the dataset lists the best odds for each fighter around the time of their fight, as well as the winner of each fight.  

 <br>
 
| **NOTE:** Some of the tables below are based on the fight data contained in the dataset described above. Although the dataset captures a large proportion of UFC events and fights since 2013, it is not exhaustive. Therefore, I will do manual searches to retrieve additional data regarding the fighters in question, to provide a more comprehensive picture this fighter-specific odds information.  |
|-----------------------------------------|

 <br>

## Over Rated Fighters

The below table lists the top 10 Over Rated fighters, in the dataset, with at least 5 UFC fights. These fighters are listed in order of Under Performance, which is simply the Average Implied Probabilities of the odds subtracted by Actual Rate of Victory. 

 <br>

```{r include = T, echo = F}
df_odds_short %>%
  gather(key = "Result", value = "NAME", Loser:Winner) -> df_odds_long

df_odds_long %>%
  dplyr::mutate(
  Was_Favorite = ifelse(
    (Favorite_was_Winner & (Result == "Winner")) | (!Favorite_was_Winner & (Result == "Loser"))
    , T
    , F
  )
) -> df_odds_long

df_odds_long %>%
  dplyr::mutate(
    Implied_Probability = ifelse(
      Was_Favorite
      , Favorite_Probability
      , Underdog_Probability
    )
    , Adjusted_Implied_Probability = ifelse(
      Was_Favorite
      , Adjusted_Favorite_Probability
      , Adjusted_Underdog_Probability
    )
  ) -> df_odds_long

df_odds_long %>% dplyr::select(
  c(
    NAME
    , Event
    , Date
    , Result
    , Implied_Probability
    , Adjusted_Implied_Probability
  )
) -> df_odds_long

df_odds_long %>%
  dplyr::mutate(
    Won = ifelse(Result == "Winner", T, F)
    , Logit_Prob = qlogis(Implied_Probability)
    , Adjusted_Logit_Prob = qlogis(Adjusted_Implied_Probability)
  ) -> df_odds_long

df_odds_long %>%
  dplyr::group_by(NAME) %>%
  dplyr::summarise(
    N_Fights = length(Won)
    , Exp_Prop = round(mean(Adjusted_Implied_Probability) * 100)
    , Win_Prop = round(mean(Won) * 100)
    , Under_Performance = -1*round(Win_Prop - Exp_Prop)
  ) -> df_odds_long_fighters

df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(desc(Under_Performance)) %>%
  head(10) -> df_top_under_perform

kable(df_top_under_perform
      , caption = "Top 10 Over Rated Fighters with at least 5 Fights"
      , col.names = c(
        "Fighter Name"
        , "Number of Fights"
        , "Average Adjusted Implied Probability (%)"
        , "Actual Rate of Victory (%)"
        , "Under Performance (%)"
       )
      )  
```

 <br>
 
I find these findings (the table above) a fair bit less interesting than those for the most [Under Rated fighters](https://aggression-to-the-mean.github.io/AttM/for-the-people/arts/2021/02/19/Under-Rated-Fighters.html).  

For starters, the names are not as big, overall, and the fighters do not have as many fights under their belts, on average. I believe that part of the reason for this is that, by definition, Over Rated fighters have not performed as well as they should have, which means they likely have not gotten as many UFC fights or been recognized as much as their peers.  

Perhaps an even more important thing to point out is that there are a few big names (Gustafsson, Evans, Hendricks, and Silva) that are on the list unfairly. Unfortunately, the dataset only captures the losing records these fighters had towards the end of their careers (i.e. after circa 2013). However, these well regarded fighters all had much more success in an earlier era (i.e. pre 2013). This era simply is not captured by the dataset.  

I am not interested in gathering all the data that is missing from the dataset. However, I did spend a bit of time analyzing Gustafsson's career. I was surprised to see how heavily he was favored in some of his match-ups.  

 <br>

# Alexander Gustafsson

The below table displays the Alexander Gustafsson fights that are included in the dataset.  

 <br>

```{r include = T, echo = F}
df_odds_long %>%
  dplyr::filter(NAME == "Alexander Gustafsson") %>%
  dplyr::select(NAME:Result, Adjusted_Implied_Probability) %>%
  dplyr::mutate(Adjusted_Implied_Probability = round(Adjusted_Implied_Probability * 100)) %>%
  dplyr::arrange(Date) -> df_gus

kable(df_gus
      , caption = sprintf("%s Fights Included in the Dataset", unique(df_gus$NAME))
      , col.names = c(
        "Fighter Name"
        , "Event"
        , "Date"
        , "Result"
        , "Adjusted Implied Probability (%)"
       )
      )
```

 <br>

I decided to focus on [Gustafsson's fights](https://www.ufc.com/athlete/alexander-gustafsson) from the first Jon Jones contest onward, since I believe it tells an interesting story.  Below are the two fights in that period of time which are not in the dataset:  

| Date  | Opponent | Result | Decimal Odds |  Opponent Decimal Odds | Adjusted Implied Probability (%)  | Under Performance (%) |
| :------------- | :----------: | :----------: | :-----------: | :-----------: | :-------------: | ----------: | 
|  2013-09-21   | Jones | Loser   |   7.60    |   1.14       |        13        |            13        | 
|  2014-03-08   | Manuwa | Winner  |   1.20  |   5.75        |         83        |           **-**17      | 

 <br>
 
**What is remarkable about the two fights listed above is that Gustafsson went from over a 6:1 underdog against Jones to nearly a 5:1 favorite against Manuwa.**

Anyway, by taking a weighted average of (i) the average Under Performance for the 2 fights above (i.e.-2%) and (ii) the average Under Performance of the 7 fights in the dataset (i.e. 34%), we get around 26%. Therefore:  

**In Gustafsson's last 9 UFC fights, from his first Jones fight until retirement, he has underperformed, relative to the odds, by about 26%.**   

I wondered if that surprisingly close first fight against Jones swung the odds in Gustafsson's favor in subsequent fights. However, if you look at [Gustafsson's odds](https://www.bestfightodds.com/fighters/Alexander-Gustafsson-1559#) leading up to the Jones fight, he was heavily favored in many of those bouts.  Therefore, it is unclear that Gustafsson's favorable odds in his past 8 fights were primarily caused by his epic performance in the first Jones bout.  

Anyway, Gustafsson's odds profile is quite strange. Overall, he seems to have been either a substantial favorite or a substantial underdog. He seems to have lost every fight for which he was the underdog (Jonesx2, Cormier, Davis). However, what makes him an Over Rated fighter in recent history are the many losses for which he was the substantial favorite.  

 <br>


## Conclusion

In this post, I discussed what made a fighter Over Rated. The results emanating from the dataset were somewhat misleading and not as interesting as those for the [Under Rated fighters](https://aggression-to-the-mean.github.io/AttM/for-the-people/arts/2021/02/19/Under-Rated-Fighters.html).  

As a result, I decided to focus on the interesting odds profile of Gustafsson. Certainly, Gustafsson appears to have been Over Rated since his first Jones contest. However, it is unclear if that close decision against Jones biased the odds in Gustafsson's favor, or if the bettors had always had such belief in his talents (preceeding the first Jones fight).   