---
title: "Fight Odds Analysis"
author: "Aggression to the Mean"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, root.dir = "/Users/Shared/AttM") 
setwd("/Users/Shared/AttM")
```

 <br>
   
## Description 

This script analyzes UFC fight odds data. 

 <br>
   
### Libraries

```{r}
library(tidyverse)
```

 <br>
   
### Examine Data

Load data.
   
```{r}
load("./Datasets/df_master.RData")
```

Get summary. 

```{r}
summary(df_master)
```

Redefine variables.

```{r}
df_master$NAME = as.factor(df_master$NAME)
df_master$Date = as.Date(df_master$Date)
df_master$Event = as.factor(df_master$Event)
df_master$City= as.factor(df_master$City)
df_master$State = as.factor(df_master$State)
df_master$Country = as.factor(df_master$Country)
df_master$FightWeightClass = as.factor(df_master$FightWeightClass)
df_master$Method = as.factor(df_master$Method)
df_master$Winner_Odds = as.numeric(df_master$Winner_Odds)
df_master$Loser_Odds = as.numeric(df_master$Loser_Odds)
df_master$fight_id = as.factor(df_master$fight_id)
df_master$Sex = as.factor(df_master$Sex)
df_master$Result = as.factor(df_master$Result)
df_master$FighterWeightClass = as.factor(df_master$FighterWeightClass)
```

Summarize again... There are infinite odds and overturned / DQ fight outcomes. These will have to be removed.  

```{r}
summary(df_master)
```

How many events does the dataset include? 

```{r}
length(unique(df_master$Event))
```
How many fights?

```{r}
length(unique(df_master$fight_id))
```

Over what time frame?

```{r}
range(sort(unique(df_master$Date)))
```

 <br>
   
### Analyse Odds

Make copy for analysis. 

```{r}
df_odds = df_master
rm(df_master)
```

Filter out controversial results and infinite odds. 

```{r}
df_odds %>%
  dplyr::filter(
    (Method != "DQ") & (Method != "Overturned")
    , is.finite(Winner_Odds)
    , is.finite(Loser_Odds)
  ) -> df_odds
```

Get rid of fighter-specifics so that we can spread the data frame. This will give us one event per row.

```{r}
df_odds %>%
  dplyr::select(-c(FighterWeight:SUBA)) %>%
  spread(Result, NAME) -> df_odds_short
```

How often were the (best) odds equal? 

```{r}
mean(df_odds$Winner_Odds == df_odds$Loser_Odds)
sum(df_odds$Winner_Odds == df_odds$Loser_Odds)
```

Filter out equal odds and identify if Favorite won the fight. 

```{r}
df_odds_short %>%
  dplyr::filter(Winner_Odds != Loser_Odds) %>%  # filter out equal odds
  dplyr::mutate(
    Favorite_was_Winner = ifelse(Winner_Odds < Loser_Odds, T, F)
    , Favorite_Unit_Profit = ifelse(Favorite_was_Winner, Winner_Odds - 1, -1)
    , Underdog_Unit_Profit = ifelse(!Favorite_was_Winner, Winner_Odds - 1, -1)
  ) -> df_odds_short
```

What was the mean unit profit (i.e. ROI) if one bet solely on the Favorite? 

```{r}
mean(df_odds_short$Favorite_Unit_Profit)
```

What was the mean unit profit if one bet solely on the Underdog?

```{r}
mean(df_odds_short$Underdog_Unit_Profit)
```

What proportion of the time does the Favorite win? 

```{r}
mean(df_odds_short$Favorite_was_Winner)
```

Calculate implied probability of each fight based on odds. 

```{r}
df_odds_short %>% dplyr::mutate(
  Favorite_Probability = ifelse(Favorite_was_Winner, 1/Winner_Odds, 1/Loser_Odds)
  , Underdog_Probability = ifelse(!Favorite_was_Winner,  1/Winner_Odds, 1/Loser_Odds)
) -> df_odds_short
```

Calculate overround for each fight.   

NOTE: these odds are the best available odds for each fight / fighter. Therefore, this is not overround in the traditional sense (looking at one particular odds maker).

```{r}
df_odds_short %>%
  dplyr::mutate(
    Total_Probability = Favorite_Probability + Underdog_Probability
    , Overround = Total_Probability - 1
  ) -> df_odds_short
```

There is very little overround. This is because we are picking the best odds for each fight / fighter. By picking the best odds, we are  counteracting the built-in overround of any particular odds-maker (typically around 5% as a rough estimate).   

```{r}
mean(df_odds_short$Overround)
mean(df_odds_short$Total_Probability)
```

 <br>
   
#### Odds performance

Add year as variable. 

```{r}
df_odds_short %>%
  dplyr::mutate(
    Year = format(Date,"%Y")
  ) -> df_odds_short
```

Create function to graphically assess over performance as a function of several variables.  These are not inferential analyses but are instead meant to visualize the data to observe trends for further analysis. 

```{r}
gauge_over_performance = function(num_bin = 10, min_bin_size = 30, variable = NULL) {

  # get bins for Favorite
  df_odds_short$Favorite_Probability_Bin = cut(df_odds_short$Favorite_Probability, num_bin)
  # get bins for Underdog
  df_odds_short$Underdog_Probability_Bin = cut(df_odds_short$Underdog_Probability, num_bin)

  if (is.null(variable)) {
    # check over/under performance for Favorites
    df_odds_short %>%
      dplyr::group_by(Favorite_Probability_Bin) %>%
      dplyr::summarise(
        Prop_of_Victory = mean(Favorite_was_Winner)
        , Size_of_Bin = length(Favorite_was_Winner)
        , ROI = mean(Favorite_Unit_Profit)
      ) -> fav_perf
  } else {

    # create dummy variable for function
    df_odds_short$Dummy = df_odds_short[
      ,which(colnames(df_odds_short) == sprintf("%s", variable))
    ]

    # check over/under performance for Favorites
    df_odds_short %>%
      dplyr::group_by(Favorite_Probability_Bin, Dummy) %>%
      dplyr::summarise(
        Prop_of_Victory = mean(Favorite_was_Winner)
        , Size_of_Bin = length(Favorite_was_Winner)
        , ROI = mean(Favorite_Unit_Profit)
      ) -> fav_perf
  }

  # extract bins
  fav_labs <- as.character(fav_perf$Favorite_Probability_Bin)
  fav_bins = as.data.frame(
    cbind(
      lower = as.numeric( sub("\\((.+),.*", "\\1", fav_labs) )
      , upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", fav_labs) )
    )
  )
  # get value in middle of bin
  fav_bins %>% dplyr::mutate(mid_bin = (lower + upper)/2 ) -> fav_bins
  # add mid bin column
  fav_perf$Mid_Bin = fav_bins$mid_bin
  # add Over performance column
  fav_perf %>% dplyr::mutate(Over_Performance = Prop_of_Victory - Mid_Bin) -> fav_perf


  if (is.null(variable)) {

    # plot over/under performance
    fav_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y=Over_Performance * 100))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("Over Performance (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Favorites")->gg
    print(gg)

    # plot over/under performance
    fav_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin * 100, y=Prop_of_Victory*100))+
      geom_point()+
      geom_smooth(se=F)+
      ylab("Probability of Victory (%)")+
      xlab("Expected Probability (%)")+
      geom_abline(slope=1, intercept=0, linetype = "dotted")+
      ggtitle("Favorites")->gg
    print(gg)

    # plot ROI - only real difference is scale along y axis
    fav_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y= ROI* 100))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("ROI (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Favorites") -> gg
    print(gg)

  } else {
    # plot over/under performance
    fav_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y=Over_Performance * 100, group=Dummy, colour = Dummy))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("Over Performance (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Favorites")+
      labs(color=sprintf("%s", variable)) -> gg
    print(gg)

    # plot ROI - only real difference is scale along y axis
    fav_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y= ROI* 100, group=Dummy, colour = Dummy))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("ROI (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Favorites")+
      labs(color=sprintf("%s", variable)) -> gg
    print(gg)
  }


  if (is.null(variable)) {

    # check over/under performance for Underdogs
    df_odds_short %>%
      dplyr::group_by(Underdog_Probability_Bin) %>%
      dplyr::summarise(
        Prop_of_Victory = mean(!Favorite_was_Winner)
        , Size_of_Bin = length(!Favorite_was_Winner)
        , ROI = mean(Underdog_Unit_Profit)
      ) -> under_perf

  } else {

    # check over/under performance for Underdogs
    df_odds_short %>%
      dplyr::group_by(Underdog_Probability_Bin, Dummy) %>%
      dplyr::summarise(
        Prop_of_Victory = mean(!Favorite_was_Winner)
        , Size_of_Bin = length(!Favorite_was_Winner)
        , ROI = mean(Underdog_Unit_Profit)
      ) -> under_perf
  }

  # extract bins
  under_labs <- as.character(under_perf$Underdog_Probability_Bin)
  under_bins = as.data.frame(
    cbind(
      lower = as.numeric( sub("\\((.+),.*", "\\1", under_labs) )
      , upper = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", under_labs) )
    )
  )
  # get value in middle of bin
  under_bins %>% dplyr::mutate(mid_bin = (lower + upper)/2 ) -> under_bins
  # add mid bin column
  under_perf$Mid_Bin = under_bins$mid_bin
  # add Over performance column
  under_perf %>% dplyr::mutate(Over_Performance = Prop_of_Victory - Mid_Bin) -> under_perf


  if (is.null(variable)) {
    # plot over/under performance
    under_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y=Over_Performance * 100))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("Over Performance (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Underdogs")->gg
    print(gg)

    # plot over/under performance
    under_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin * 100, y=Prop_of_Victory*100))+
      geom_point()+
      geom_smooth(se=F)+
      ylab("Probability of Victory (%)")+
      xlab("Expected Probability (%)")+
      geom_abline(slope=1, intercept=0, linetype = "dotted")+
      ggtitle("Underdogs")->gg
    print(gg)

    under_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y=ROI * 100))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("ROI (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Underdogs")-> gg
    print(gg)

  } else {

    # plot over/under performance
    under_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y=Over_Performance * 100, group=Dummy, colour = Dummy))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("Over Performance (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Underdogs")+
      labs(color=sprintf("%s", variable)) -> gg
    print(gg)

    under_perf %>%
      dplyr::filter(Size_of_Bin >= min_bin_size) %>%
      ggplot(aes(x=Mid_Bin*100, y=ROI * 100, group=Dummy, colour = Dummy))+
      geom_point()+
      geom_smooth(se=F)+
      geom_hline(yintercept = 0, linetype = "dotted")+
      ylab("ROI (%)")+
      xlab("Expected Probability (%)")+
      ggtitle("Underdogs")+
      labs(color=sprintf("%s", variable)) -> gg
    print(gg)
  }
  
  # process to return()
  under_perf$Is_Fav = F
  under_perf %>%
    rename(Probability_Bin = Underdog_Probability_Bin) -> under_perf
  
  fav_perf$Is_Fav = T
  fav_perf %>%
    rename(Probability_Bin = Favorite_Probability_Bin) -> fav_perf
  
  return(rbind(fav_perf, under_perf))

}
```

Look at how expected performance predicts over performance. 

```{r}
gauge_over_performance(num_bin = 10, min_bin_size = 100, variable = NULL)
```

Is there any stability across years? Need to reduce minimum bin size to get estimates. As a result, estimates will be more noisy. 

```{r}
gauge_over_performance(num_bin = 10, min_bin_size = 30, variable = "Year")
```

Does the method of victory affect the relationship between odds and outcome? Reduce number of bins (compared to Year comparison above) to stabilize estimates. Graphs do not tell whole story due to number of data points available across bins. 

```{r}
odds_perf_by_method = gauge_over_performance(num_bin = 5, min_bin_size = 30, variable = "Method")
print(odds_perf_by_method)
```

How does fight finishing method vary with implied probability of vegas odds? 

```{r}
odds_perf_by_method %>%
  dplyr::filter(Is_Fav == T) %>%
  ggplot(aes(x=Mid_Bin, y=Size_of_Bin, group = Dummy, color = Dummy))+
  geom_point()+
  geom_smooth(se=F)+
  ylab("Count")+
  xlab("Implied Probability (%)")+
  ggtitle("Favorites")+
  labs(color="Method")

odds_perf_by_method %>%
  dplyr::filter(Is_Fav == F) %>%
  ggplot(aes(x=Mid_Bin, y=Size_of_Bin, group = Dummy, color = Dummy))+
  geom_point()+
  geom_smooth(se=F)+
  ylab("Count")+
  xlab("Implied Probability (%)")+
  ggtitle("Underdogs")+
  labs(color="Method")
```

Calculate the proportion of fights that end by various methods as a function of implied probability of fight odds.

```{r}
odds_perf_by_method %>%
  group_by(Is_Fav, Mid_Bin) %>%
  summarise(Total_Count = sum(Size_of_Bin)) -> total_count

odds_perf_by_method %>%
  group_by(Is_Fav, Mid_Bin, Dummy) %>%
  summarise(Count= Size_of_Bin) -> single_count

method_count_by_odds = merge(single_count, total_count)
method_count_by_odds %>%
  dplyr::mutate(Method_Prop = Count / Total_Count ) -> method_count_by_odds

method_count_by_odds %>%
  dplyr::filter(Is_Fav == T) %>%
  ggplot(aes(x=Mid_Bin, y=Method_Prop, group = Dummy, color=Dummy))+
  geom_point()+
  geom_smooth(se=F)+
  ylab("Proportion")+
  xlab("Implied Probability (%)")+
  ggtitle("Favorites")+
  labs(color="Method")

method_count_by_odds %>%
  dplyr::filter(Is_Fav == F) %>%
  ggplot(aes(x=Mid_Bin, y=Method_Prop, group = Dummy, color=Dummy))+
  geom_point()+
  geom_smooth(se=F)+
  ylab("Proportion")+
  xlab("Implied Probability (%)")+
  ggtitle("Underdogs")+
  labs(color="Method")
```

 <br>
   

#### Fighter Odds 

Get rid of useless columns. 

```{r}
df_odds %>% dplyr::select(
  c(
    NAME
    , Event
    , Date
    , Result
    , Winner_Odds
    , Loser_Odds
  )
) -> df_odds_long
```

Summarize data.

```{r}
summary(df_odds_long)
```

Add Fighter Odds column.

```{r}
df_odds_long %>%
  dplyr::mutate(
    Fighter_Odds = ifelse(Result == "Winner", Winner_Odds, Loser_Odds)
  ) -> df_odds_long
```

Add Implied Probability column.  

```{r}
df_odds_long %>%
  dplyr::mutate(
    Implied_Probability = 1/Fighter_Odds
    , Won = ifelse(Result == "Winner", T, F)
    , Logit_Prob = qlogis(Implied_Probability)
  ) -> df_odds_long
```

Get performance and odds.

```{r}
df_odds_long %>%
  dplyr::group_by(NAME) %>%
  dplyr::summarise(
    Exp_Prop = mean(Implied_Probability)
    , Logit_Exp_Prop = mean(Logit_Prob)
    , Win_Prop = mean(Won)
    , N_Fights = length(Won)
    , Over_Performance = Win_Prop - Exp_Prop
    , Logit_Over = qlogis(Win_Prop) - Logit_Exp_Prop
    , Back_Trans_Exp = plogis(Logit_Exp_Prop)
  ) -> df_odds_long_fighters
```

Top 10 over-performers with at least 5 fights where number of fights is simply number available in the dataset (see above). 

```{r}
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(desc(Over_Performance))
# now with logit
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(desc(Logit_Over))
```

Top 10 under performers with at least 5 fights. 

```{r}
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(Over_Performance)
# with logit
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(Logit_Over)
```

Most highly favorited fighters with at least 5 fights

```{r}
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(desc(Exp_Prop))
# with logit
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(desc(Logit_Exp_Prop))
```

Most undervalued fighters with at least 5 fights. 

```{r}
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(Exp_Prop)
# with logit
df_odds_long_fighters %>%
  dplyr::filter(N_Fights >= 5) %>%
  dplyr::arrange(Logit_Exp_Prop)
```

Examine odds for specific fighters.   

```{r}
# Israel Adesanya
df_odds_long_fighters %>% dplyr::filter(NAME == "Israel Adesanya")

# Anthony Smith
df_odds_long_fighters %>% dplyr::filter(NAME == "Anthony Smith")
```

